VIPRE: Virtualized Intermittently-Powered Runtime Environment

### 2017-07-10 paper/intro.tex
All references of the intro cleaned and missing added
Paragraph 4 ready

Advances in processor efficiency along with the development of energy-harvesting power systems has created a new category of devices that require neither a battery nor a tethered power supply~\cite{prasad_comst_2014,lucia_snapl_2017,soyata_csm_2016}. These devices operate entirely using energy extracted from their environment, such as RF energy~\cite{rf_powered_computing_gollakota_2014}, photovoltaic components~\cite{margolies_infocom_2016,margolies_tosn_2016}, and vibration~\cite{gorlatova_sigmetrics_2014}. Incorporating compute, storage, sensing, and communication hardware, such energy-independent devices are a promising underlying technology for applications in emerging areas such as Internet of Things (IoT)~\cite{ku_cst_2016}, in-body~\cite{nadeau_naturebio_2017} and on-body~\cite{bandodkar_electroanalysis_2015} medical systems, and energy-harvesting nano-satellites~\cite{kicksat}.

While promising, energy-harvesting devices present a unique design challenge in
that they operate only {\em intermittently} when energy is available~\cite{hicks_isca_2017,lucia_snapl_2017}. An
energy-harvesting device buffers energy from its environment, e.g., in a
capacitor~\cite[Fig. 3]{gorlatova_tmc_2013},~\cite[Fig. 1]{gunduz_commag_2014}. When a threshold quantity of energy accumulates, the device begins operating. Harvestable energy sources typically produce relatively very little power compared to a platform's operating power level---compare for instance~\cite[Table III and V]{prasad_comst_2014} with~\cite[Table I]{carrano_cst_2014}. A devices operates for a brief period of time and when buffered energy is depleted, shuts down and begins recharging to later operate again. As an example, inter-task time may even exceed 20\,s for wireless energy-harvesting battery-less vivo digestive tract monitoring system~\cite[Fig. 3c]{nadeau_naturebio_2017}. Charge and discharge times vary by device and some of them~\cite{wisp} may fail $\approx$10 to $\approx$100 times per second~\cite[Fig. 1]{tan_infocom_2016},~\cite[Fig. 1]{mementos}.

\todo{Put exact reference for "tictpl"}{Brandon}

Software in an energy-harvesting system operates according to the {\em
intermittent execution model}~\cite[Sec. 2]{dino}~\cite[Fig. 2]{lucia_snapl_2017}, with bursts of execution that are interrupted by failure periods. When power fails, a device loses its volatile state, i.e., registers, stack, SRAM, and retains its non-volatile state, i.e., FRAM, Flash. While capturing periodic checkpoints~\cite{mementos,quickrecall} and sleep scheduling~\cite{dewdrop,hibernus,hibernusplusplus} help to preserve execution progress, failures can leave volatile and non-volatile state inconsistent, leading to unrecoverable failures~\cite{mspcdino,edb}.

@@

New programming and execution models~\cite{dino,ratchet,chain,alpaca} and
architecture mechanisms~\cite{hicks_isca_2017,idetic,nvp} help avoid data
inconsistency. New architectures require costly hardware changes and are
inapplicable to today's systems. New programming models and compilers force
programmers to use new memory models~\cite{chain,ratchet}, which may inhibit
adoption. New memory models may also increase the amount of non-volatile
memory accesses by requiring non-volatile versioning~\cite{chain} or completely
precluding the use of volatile memory~\cite{ratchet}. Increasing pressure on
non-volatile memory or precluding the use of volatile memory may degrade
efficiency; e.g., FRAM has a higher access energy and latency than
SRAM~\cite{alpaca,wisp}. Commercially available devices are often provisioned
with much less SRAM than FRAM~\cite{wolverine,wisp} and effectively leveraging
the relatively high efficiency of SRAM is a key challenge for task-based
intermittent systems.

\TODO{concrete cost multiples here?}

%Task sizing is hard -- how to chop without risking inefficiency or non-term
Several recent programming efforts~\cite{alpaca,chain} proposed {\em
task-based} programming models, that require the programmer to statically
decompose their program into tasks. A task is like a function with no callers
that can include arbitrary computation and, upon completion, is guaranteed to
have executed {\em atomically} and to have logically {\em committed} memory
updates. The programmer expresses the flow of control from one task to another
and task control-flow may be input-dependent. An important programming
challenge in task-based models is that the length of execution of a task is
limited by the total amount of energy that a device can buffer. Assuming that
input power is negligible compared to operating power, a task will never be
able to complete if its execution consumes more energy than the system can
buffer. Transitioning from one task to the next imposes a cost and excessively
small tasks may be inefficient. Appropriately sizing tasks to ensure progress
and high performance is a key challenge for task-based intermittent systems.

>>>>

Advances in processor efficiency along with the development of energy-harvesting power systems has created a new category of devices that require neither a battery nor a tethered power supply~\cite{prasad_comst_2014,lucia_snapl_2017,soyata_csm_2016}. These devices operate entirely using energy extracted from their environment, such as RF energy~\cite{rf_powered_computing_gollakota_2014}, photovoltaic components~\cite{margolies_infocom_2016,margolies_tosn_2016}, and vibration~\cite{gorlatova_sigmetrics_2014}. Incorporating compute, storage, sensing, and communication hardware~\cite{wisp5,moo}, such energy-independent devices are a promising underlying technology for applications in emerging areas such as Internet of Things (IoT)~\cite{ku_cst_2016}, in-body~\cite{nadeau_naturebio_2017} and on-body~\cite{bandodkar_electroanalysis_2015} medical systems, and energy-harvesting nano-satellites~\cite{kicksat}.

While promising, energy-harvesting devices present a unique design challenge in
that they operate only {\em intermittently} when energy is available~\cite{hicks_isca_2017,lucia_snapl_2017}. An
energy-harvesting device buffers energy from its environment, e.g., in a
capacitor~\cite[Fig. 3]{gorlatova_tmc_2013},~\cite[Fig. 1]{gunduz_commag_2014}. When a threshold quantity of energy accumulates, the device begins operating. Harvestable energy sources typically produce relatively very little power compared to a platform's operating power level---compare for instance~\cite[Table III and V]{prasad_comst_2014} with~\cite[Table I]{carrano_cst_2014}. A devices operates for a brief period of time and when buffered energy is depleted, shuts down and begins recharging to operate again later. As an example, inter-task time may even exceed 20\,s for wireless energy-harvesting battery-less vivo digestive tract monitoring system~\cite[Fig. 3c]{nadeau_naturebio_2017}. Charge and discharge times vary by device and some of them~\cite{wisp} may fail $\approx$10 to $\approx$100 times per second~\cite[Fig. 1]{tan_infocom_2016},~\cite[Fig. 1]{mementos},~\cite[Fig. 3]{nvp}.

\todo{Put exact reference for "tictpl"}{Brandon}
\todo{Check if the list of authors for Alpaca is correct}{Brandon}

Software in an energy-harvesting system operates according to the {\em
intermittent execution model}~\cite[Sec. 2]{dino}~\cite[Fig. 2]{lucia_snapl_2017}, with bursts of execution interrupted by failure periods. When power fails, a device loses its volatile state, i.e., registers, stack, SRAM, and retains its non-volatile state, i.e., FRAM, Flash. While capturing periodic checkpoints~\cite{mementos,quickrecall} and sleep scheduling~\cite{dewdrop,hibernus,hibernusplusplus} help to preserve execution progress, failures can leave volatile and non-volatile state inconsistent, leading to unrecoverable failures~\cite{mspcdino,edb}.

@@

New programming and execution models~\cite{dino,ratchet,chain,alpaca} and
architecture mechanisms~\cite{hicks_isca_2017,idetic,nvp} for intermittently-powered devices help avoid data inconsistency. Nonetheless, both approaches---models and architectures---posses critical limitations. New architectures require costly hardware changes and are inapplicable to today's systems. For instance, they require dedicated idempotency violation control hardware buffers~\cite[Fig. 3]{hicks_isca_2017} or dedicated on-processor logic like on demand register backup~\cite[Fig. 9]{nvp}. Then, new programming models and compilers force programmers to use new memory models~\cite{chain,ratchet}, which may inhibit widespread adoption by programmers accustomed to classical (C) program writing. New memory models may also increase the amount of non-volatile memory accesses by requiring non-volatile versioning~\cite{chain} or completely precluding the use of volatile memory~\cite{ratchet}. Increasing pressure on non-volatile memory or precluding the use of volatile memory may degrade efficiency; e.g., FRAM has a higher access energy and latency than SRAM~\cite[Fig. 4]{nvp}---which will be also shown quantitatively in this paper. Commercially available devices are often provisioned with much less SRAM than FRAM, e.g. up to 64 FRAM/SRAM ratio of~\cite{wolverine}, and effectively leveraging the relatively high efficiency of SRAM is a key challenge for task-based intermittent systems.

%Task sizing is hard -- how to chop without risking inefficiency or non-term

Several recent programming efforts~\cite{alpaca,chain} proposed {\em
task-based} programming models, that require the programmer to statically
decompose their program into tasks. A task is like a function with no callers
that can include arbitrary computation and, upon completion, is guaranteed to
have executed {\em atomically} and to have logically {\em committed} memory
updates. The programmer expresses the flow of control from one task to another
and task control-flow may be input-dependent. An important programming
challenge in task-based models is that the length of execution of a task is
limited by the total amount of energy that a device can buffer. Assuming that
input power is negligible compared to operating power, a task will never be
able to complete if its execution consumes more energy than the system can
buffer. Transitioning from one task to the next imposes a cost and excessively
small tasks may be inefficient. Appropriately sizing tasks to ensure progress
and high performance is a key challenge for task-based intermittent systems.
